---
title: "niche overlap"
output: html_document
date: "2024-07-26"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reading in libs
```{r}
library(ggplot2)
library(phyloseq)
library(tidyverse)
library(here)
library(vegan)
library(microViz)
library(ggpubr)
library(spaa)
```

Read in our data 
```{r}
physeq_rra <- readRDS("banff_diet_phyloseq_cleaned_Dec032024.rds") 


physeq_rra_core <- subset_samples(physeq_rra, (Species == "Bison" # filter for bison, elk, sheep
                                                      | Species == "BighornSheep" | 
                                                        Species == "Elk"))

sample_data(physeq_rra)$Season = factor(sample_data(physeq_rra)$Season, 
                                         levels = c("Spring", "Summer", "Winter", "Fall"), 
                                         labels = c("Spring", "Summer", "Winter", "Winter"))
sample_data(physeq_rra_core)$Season = factor(sample_data(physeq_rra_core)$Season, 
                                         levels = c("Spring", "Summer", "Winter", "Fall"), 
                                         labels = c("Spring", "Summer", "Winter", "Winter"))


sample_data(physeq_rra)$Species = factor(sample_data(physeq_rra)$Species, 
                                         levels = c("Bison", "BighornSheep",  "Elk", "Goat",  "Deer", "Moose" ), 
                                         labels = c("Bison", "Bighorn Sheep", "Elk",
                                                     "Mountain Goat",  "Deer Spp.", "Moose"))


sample_data(physeq_rra)$Season = factor(sample_data(physeq_rra)$Season, levels = c("Winter", "Summer", "Spring")) 


```



DIFFERENCES IN DIET DUE TO MORPHOLOGY 
```{r}
sample_data(physeq_rra)$feeding_mode <- if_else(sample_data(physeq_rra)$Species == "Bison" | 
                                                  sample_data(physeq_rra)$Species == "BighornSheep", "Grazer", 
                                if_else(sample_data(physeq_rra)$Species== "Deer", "Browser", "Mixed"))

feeding_mode_dat <- subset_samples(physeq_rra, (Species != "Moose" & Species != "Goat"))



# group species by functional feeding more and see if permanova differentiates them... use spicy pairwise comparisons too!! 
diet_matrix <- t(otu_table(feeding_mode_dat)) # get our community matrix
covar_table <- data.frame(sample_data(feeding_mode_dat)) 
# run adonis test... community data matrix ~ covariate (rows in community matrix match rows in covar data frame)
table(covar_table$feeding_mode)
ffmode_mod <- adonis2(diet_matrix ~ feeding_mode, data = covar_table, permutations = 999, 
                method = "bray",
    parallel = 15, na.action = na.fail)

ffmode_mod


# group species by functional feeding more and see if permanova differentiates them... use spicy pairwise comparisons too!! 
graz_browse <- subset_samples(feeding_mode_dat, (feeding_mode == "Grazer" # filter for bison, elk, sheep
                                                      | feeding_mode == "Browser"))

diet_matrix <- t(otu_table(graz_browse)) # get our community matrix
covar_table <- data.frame(sample_data(graz_browse)) 
# run adonis test... community data matrix ~ covariate (rows in community matrix match rows in covar data frame)
table(covar_table$feeding_mode)
ffmode_mod <- adonis2(diet_matrix ~ feeding_mode, data = covar_table, permutations = 999, 
                method = "bray",
    parallel = 15, na.action = na.fail)

ffmode_mod

## graz graz! 
graz_browse <- subset_samples(feeding_mode_dat, (feeding_mode == "Grazer" # filter for bison, elk, sheep
                                                      | feeding_mode == "Mixed"))

diet_matrix <- t(otu_table(graz_browse)) # get our community matrix
covar_table <- data.frame(sample_data(graz_browse)) 
# run adonis test... community data matrix ~ covariate (rows in community matrix match rows in covar data frame)
table(covar_table$feeding_mode)
ffmode_mod <- adonis2(diet_matrix ~ feeding_mode, data = covar_table, permutations = 999, 
                method = "bray",
    parallel = 15, na.action = na.fail)

ffmode_mod
p.adjust(ffmode_mod$`Pr(>F)`, "holm", n = 3) 


graz_browse <- subset_samples(feeding_mode_dat, (feeding_mode == "Browser" # filter for bison, elk, sheep
                                                      | feeding_mode == "Mixed"))

diet_matrix <- t(otu_table(graz_browse)) # get our community matrix
covar_table <- data.frame(sample_data(graz_browse)) 
# run adonis test... community data matrix ~ covariate (rows in community matrix match rows in covar data frame)
table(covar_table$feeding_mode)
ffmode_mod <- adonis2(diet_matrix ~ feeding_mode, data = covar_table, permutations = 999, 
                method = "bray",
    parallel = 15, na.action = na.fail)

ffmode_mod

p.adjust(ffmode_mod$`Pr(>F)`, "holm", n = 3) 
```

BODY SIZE DIFFERENCES? 
```{r}
######## PAIRWISE COMPARISONS 

# bison sheep 
bis_bhs <- subset_samples(physeq_rra, Species == "Bison" | Species == "Bighorn Sheep")
diet_matrix <- t(otu_table(bis_bhs)) # get our community matrix
covar_table <- data.frame(sample_data(bis_bhs))
bis_bhs_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)

# bison elk 
bis_elk <- subset_samples(physeq_rra, Species == "Bison" | (Species == "Elk" & Long < -115.4105))
diet_matrix <- t(otu_table(bis_elk)) # get our community matrix
covar_table <- data.frame(sample_data(bis_elk))
bis_elk_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)

# bison deer
bis_deer  <- subset_samples(physeq_rra, Species == "Bison" | Species == "Deer Spp.")
diet_matrix <- t(otu_table(bis_deer)) # get our community matrix
covar_table <- data.frame(sample_data(bis_deer))
bis_deer_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bis_deer_pair


# bighorn slk 
dat  <- subset_samples(physeq_rra, Species == "Bighorn Sheep" | Species == "Elk")
diet_matrix <- t(otu_table(dat)) # get our community matrix
covar_table <- data.frame(sample_data(dat))
bhs_elk_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bhs_elk_pair


# bighorn deer 
dat  <- subset_samples(physeq_rra, Species == "Bighorn Sheep" | Species == "Deer Spp.")
diet_matrix <- t(otu_table(dat)) # get our community matrix
covar_table <- data.frame(sample_data(dat))
bhs_deer_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bhs_deer_pair


# elk - deer 
dat  <- subset_samples(physeq_rra, Species == "Elk" | Species == "Deer Spp.")
diet_matrix <- t(otu_table(dat)) # get our community matrix
covar_table <- data.frame(sample_data(dat))
elk_deer_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
elk_deer_pair



# get our adjusted p values
pairwise.species <- data.frame(species = c("bis_bhs", "bis_elk", "bis_deer", 
                                           "bhs_elk", "bhs_deer",
                                           "elk_deer"), 
                              rsq = c(bis_bhs_pair$R2[1],
                                        bis_elk_pair$R2[1], 
                                        bis_deer_pair$R2[1],
                                      #
                                      bhs_elk_pair$R2[1],
                                        bhs_deer_pair$R2[1],
                                      #
                                        elk_deer_pair$R2[1]),
                                      #
                                  pseudoF = c(bis_bhs_pair$F[1],
                                        bis_elk_pair$F[1], 
                                        bis_deer_pair$F[1],
                                      #
                                      bhs_elk_pair$F[1],
                                        bhs_deer_pair$F[1],
                                      #
                                        elk_deer_pair$F[1]),
                                  pval = c(bis_bhs_pair$`Pr(>F)`[1],
                                        bis_elk_pair$`Pr(>F)`[1], 
                                        bis_deer_pair$`Pr(>F)`[1],
                                      #
                                      bhs_elk_pair$`Pr(>F)`[1],
                                        bhs_deer_pair$`Pr(>F)`[1],
                                      #
                                        elk_deer_pair$`Pr(>F)`[1]),
                                  
                                body_diff = c(550, 375, 545, 175, 5, 170))


pairwise.species$holm <-   p.adjust(pairwise.species$pval, "holm") 


write.csv(pairwise.species, "Outputs/pairwise_comparisons_across_all_species_dec082024.csv")


pairwise.species <- read.csv("Outputs/pairwise_comparisons_across_all_species_dec082024.csv")

####### BODY SIZE DIFFERENCES USING MANTEL TEST! 
# Split species names
pairwise.species_split <- tidyr::separate(pairwise.species, species, into=c("sp1","sp2"), sep="_") 
pairwise.species_split <- rbind(pairwise.species_split, data.frame(X = 7:10, sp1 = c("bis", "bhs", "elk", "deer"),sp2= c("bis", "bhs", "elk", "deer"), 
                                                                   rsq = NA, pseudoF = NA, pval = NA, body_diff  = NA,holm = NA))

# Make square matrix of species × species
species <- unique(c(pairwise.species_split$sp1, pairwise.species_split$sp2))
rsq_mat <- matrix(NA, nrow=length(species), ncol=length(species),
                  dimnames=list(species, species))
size_mat <- matrix(NA, nrow=length(species), ncol=length(species),
                  dimnames=list(species, species))
# Fill matrix
for(i in 1:nrow(pairwise.species_split)) {
  rsq_mat[pairwise.species_split$sp1[i], pairwise.species_split$sp2[i]] <- pairwise.species_split$rsq[i]
  rsq_mat[pairwise.species_split$sp2[i], pairwise.species_split$sp1[i]] <- pairwise.species_split$rsq[i]  # make symmetric
}
diag(rsq_mat) <- 0
# Fill matrix
for(i in 1:nrow(pairwise.species_split)) {
  size_mat[pairwise.species_split$sp1[i], pairwise.species_split$sp2[i]] <- pairwise.species_split$body_diff[i]
  size_mat[pairwise.species_split$sp2[i], pairwise.species_split$sp1[i]] <- pairwise.species_split$body_diff[i]  # make symmetric
}
diag(size_mat) <- 0

# converting to distance objects!
diss_dist <- as.dist(rsq_mat)
size_dist <- as.dist(size_mat)

# mantel test! 
mantel_res <- mantel(diss_dist, size_dist, method = "pearson", permutations = 999)
print(mantel_res)


```



Adonis test to examine if there are seasonal differences in diets of elk/bison/sheep. Followed by some nice pairwise comaprisons
```{r}
## all species-season! 
diet_matrix <- t(otu_table(physeq_rra_core)) # get our community matrix
covar_table <- data.frame(sample_data(physeq_rra_core))
class(covar_table)
# run adonis test... community data matrix ~ covariate (rows in community matrix match rows in covar data frame)
season_mod <- adonis2(diet_matrix ~ Species*Season, data = covar_table, permutations = 999, 
                method = "bray",
    parallel = 15, na.action = na.fail)

season_mod

#rm(season_mod)

# WITHIN season differences 

sp <- subset_samples(physeq_rra_core, Season == "Spring")
diet_matrix <- t(otu_table(sp)) # get our community matrix
covar_table <- data.frame(sample_data(sp))
sp_adonis <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
sp_adonis

# SPRING SIMPER analysis
sp_simper_result <- simper(diet_matrix, covar_table$Species, permutations = 999)
summary(sp_simper_result)

su <- subset_samples(physeq_rra_core, Season == "Summer")
diet_matrix <- t(otu_table(su)) # get our community matrix
covar_table <- data.frame(sample_data(su))
su_adonis <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
su_adonis
su_simper_result <- simper(diet_matrix, covar_table$Species, permutations = 999)
summary(su_simper_result)



w <- subset_samples(physeq_rra_core, Season == "Winter" | Season == "Fall")
diet_matrix <- t(otu_table(w)) # get our community matrix
covar_table <- data.frame(sample_data(w))
w_adonis <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
w_adonis
w_simper_result <- simper(diet_matrix, covar_table$Species, permutations = 999)
summary(w_simper_result)

############# PAIRWISE ADONIS ##########################
# use adonis to compare species-season.... 
#################### SEASON SPECIFIC COMPARISONS 

# elk and bison winter
bis_elk <- subset_samples(physeq_rra_core, Season == "Winter" & (Species == "Bison" | Species == "Elk"))
diet_matrix <- t(otu_table(bis_elk)) # get our community matrix
covar_table <- data.frame(sample_data(bis_elk))
bis_elk_w_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bis_elk_w_pair

# bhs and bison winter
bis_bhs <- subset_samples(physeq_rra_core, Season == "Winter" & (Species == "Bison" | Species == "BighornSheep"))
diet_matrix <- t(otu_table(bis_bhs)) # get our community matrix
covar_table <- data.frame(sample_data(bis_bhs))
bis_bhs_w_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bis_bhs_w_pair

# elk and bison spring
bis_elk <- subset_samples(physeq_rra_core, Season == "Spring" & (Species == "Bison" | Species == "Elk"))
diet_matrix <- t(otu_table(bis_elk)) # get our community matrix
covar_table <- data.frame(sample_data(bis_elk))
bis_elk_sp_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bis_elk_sp_pair

# bhs and bison spring
bis_bhs <- subset_samples(physeq_rra_core, Season == "Spring" & (Species == "Bison" | Species == "BighornSheep"))
diet_matrix <- t(otu_table(bis_bhs)) # get our community matrix
covar_table <- data.frame(sample_data(bis_bhs))
bis_bhs_sp_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bis_bhs_sp_pair

# elk and bison summer
bis_elk <- subset_samples(physeq_rra_core, Season == "Summer" & (Species == "Bison" | Species == "Elk"))
diet_matrix <- t(otu_table(bis_elk)) # get our community matrix
covar_table <- data.frame(sample_data(bis_elk))
bis_elk_su_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bis_elk_su_pair

# bhs and bison SUMMER
bis_bhs <- subset_samples(physeq_rra_core, Season == "Summer" & (Species == "Bison" | Species == "BighornSheep"))
diet_matrix <- t(otu_table(bis_bhs)) # get our community matrix
covar_table <- data.frame(sample_data(bis_bhs))
bis_bhs_su_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bis_bhs_su_pair

# bhs and elk spring
dat <- subset_samples(physeq_rra, Season == "Spring" & (Species == "Elk" | Species == "BighornSheep"))
diet_matrix <- t(otu_table(dat)) # get our community matrix
covar_table <- data.frame(sample_data(dat))
bhs_elk_sp_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bhs_elk_sp_pair

# bjs and elk summer
dat <- subset_samples(physeq_rra, Season == "Summer" & (Species == "Elk" | Species == "BighornSheep"))
diet_matrix <- t(otu_table(dat)) # get our community matrix
covar_table <- data.frame(sample_data(dat))
bhs_elk_su_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bhs_elk_su_pair


# bjs and elk winter
dat <- subset_samples(physeq_rra, Season == "Winter" & (Species == "Elk" | Species == "BighornSheep"))
diet_matrix <- t(otu_table(dat)) # get our community matrix
covar_table <- data.frame(sample_data(dat))
bhs_elk_w_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bhs_elk_w_pair


# vector of p values 
pairwise.seasonal <- data.frame(species = c("bis_elk_w_pair", "bis_bhs_w_pair", "bis_elk_sp_pair", "bis_bhs_sp_pair", 
                                   "bis_elk_su_pair", "bis_bhs_su_pair", 
                                   "bhs_elk_sp", "bhs_elk_su", "bhs_elk_w"),
                      rsq = c(bis_elk_w_pair$R2[1], 
                                  bis_bhs_w_pair$R2[1], 
                                  bis_elk_sp_pair$R2[1], 
                                  bis_bhs_sp_pair$R2[1], 
                                  bis_elk_su_pair$R2[1], 
                                  bis_bhs_su_pair$R2[1],
                              bhs_elk_sp_pair$R2[1],
                              bhs_elk_su_pair$R2[1],
                              bhs_elk_w_pair$R2[1]
                              ), 
                      pseudoF = c(bis_elk_w_pair$F[1], 
                                  bis_bhs_w_pair$F[1], 
                                  bis_elk_sp_pair$F[1], 
                                  bis_bhs_sp_pair$F[1], 
                                  bis_elk_su_pair$F[1], 
                                  bis_bhs_su_pair$F[1], 
                                
                              bhs_elk_sp_pair$F[1],
                              bhs_elk_su_pair$F[1],
                              bhs_elk_w_pair$F[1]),
                       pval = c( bis_elk_w_pair$`Pr(>F)`[1], 
                                  bis_bhs_w_pair$`Pr(>F)`[1], 
                                  bis_elk_sp_pair$`Pr(>F)`[1], 
                                  bis_bhs_sp_pair$`Pr(>F)`[1], 
                                  bis_elk_su_pair$`Pr(>F)`[1], 
                                  bis_bhs_su_pair$`Pr(>F)`[1], 
                                 
                              bhs_elk_sp_pair$`Pr(>F)`[1],
                              bhs_elk_su_pair$`Pr(>F)`[1],
                              bhs_elk_w_pair$`Pr(>F)`[1]))
pairwise.seasonal$holm <-   p.adjust(pairwise.seasonal$pval, "holm") 

write.csv(pairwise.seasonal, "Outputs/INSIDE_bisZONE_pairwise_comparisons_across_season.csv")
```



Visualizing using NMDS 
Seasonal differences visualized (bison, sheep, elk) --> ALL SEASONS 
```{r}
GP.ord <- ordinate(physeq_rra , "NMDS", "bray", trymax=300) # run ordination 

ordination_RRA_dat = plot_ordination(physeq_rra, GP.ord, color = "Species", shape = "Species", type="samples", justDF = TRUE) %>% 
  group_by(Species, Season) %>% 
  mutate(mean_nmds1 = mean(NMDS1), 
          mean_nmds2 = mean(NMDS2),
          se_nmds1 = sd(NMDS1)/sqrt(n_distinct(SampleID)), 
          se_nmds2 = sd(NMDS2)/sqrt(n_distinct(SampleID)))
  

ordination_RRA_dat$Species =factor(ordination_RRA_dat$Species, levels = c("Bison",  
                                                   "Bighorn Sheep",    "Elk",  "Goat",  "Deer",  "Moose"),
                            labels = c("Bison",  "Bighorn Sheep", "Elk", "Mountain Goat", "Deer",  "Moose"))

ordination_RRA <- ggplot(ordination_RRA_dat, aes(x = NMDS1, y = NMDS2*-1, 
                                                 color = Species, linetype = Species, shape = Species)) +
  geom_point(size = 1.5) + 
  stat_chull(aes(fill = Species, linetype = Species), geom = "polygon", alpha = 0.2, size = 0.8) + 
  scale_color_manual(values = c("#586028","tan1",  "#00A1B7",  "grey30",  "grey50",  "grey80")) +
  scale_fill_manual(values = c("#586028","tan1",  "#00A1B7",  "grey30",  "grey50",  "grey80")) +
  scale_shape_manual(values = c(16, 16, 15, 15, 17, 17)) + 
  scale_linetype_manual(values = c("solid",  "solid", "dashed", "dashed", "dotted", "dotted")) + 
  ylab("NMDS2") + 
     theme_bw() + 
  theme(legend.position =  c(0.75, 0.23), legend.text = element_text(size = 16), legend.title = element_blank(),
        strip.text = element_text(size = 16),   
        axis.text = element_blank(), axis.ticks = element_blank(),
        axis.title = element_text(size = 16)) + 
 # geom_text(aes(label = ID_Number), vjust = -0.5) + 
  facet_wrap(.~Season, nrow = 2,  strip.position = "right") 
ordination_RRA

ggsave("Outputs/all_species_dietoord_seasons_Sept102025.jpeg", ordination_RRA, height = 8, width =6)




################## Removing deer, moose, and goats (from the figure in main text) 
GP.ord <- ordinate(physeq_rra_core , "NMDS", "bray", trymax=300) # run ordination 

ordination_RRA_dat = plot_ordination(physeq_rra_core, GP.ord, color = "Species", shape = "Species", type="samples", justDF = TRUE) %>% 
  group_by(Species, Season) %>% 
  mutate(mean_nmds1 = mean(NMDS1), 
          mean_nmds2 = mean(NMDS2),
          se_nmds1 = sd(NMDS1)/sqrt(n_distinct(SampleID)), 
          se_nmds2 = sd(NMDS2)/sqrt(n_distinct(SampleID)))

ordination_RRA_dat$Species = factor(ordination_RRA_dat$Species, levels = c("Bison", "BighornSheep", "Elk"), 
                                    labels = c("Bison", "Bighorn sheep", "Elk"))
table(ordination_RRA_dat$Species)

ordination_RRA_dat$Season = factor(ordination_RRA_dat$Season, levels = c("Winter", "Summer", "Spring"))

ordination_RRA <- ggplot(ordination_RRA_dat, aes(x = NMDS1, y = NMDS2*-1, 
                                                 color = Species, linetype = Species, 
                                                 shape = Species)) +
  geom_point(size = 1.5) + 
  stat_chull(aes(fill = Species, linetype = Species), geom = "polygon", alpha = 0.5, size = 0.8) + 
  scale_fill_manual(values = c("#586028","tan1",  "#00A1B7")) + 
  scale_color_manual(values = c("#586028","tan1",  "#00A1B7")) + 
  scale_shape_manual(values = c(16, 15,  17)) + 
  scale_linetype_manual(values = c("solid", "dashed",  "dotted")) + 
  ylab("NMDS2") + 
  theme_bw() + 
  theme(legend.position ="none", 
        strip.text = element_text(size = 16),   
        axis.text = element_blank(), axis.ticks = element_blank(),
        axis.title = element_text(size = 16)) + 

    facet_wrap(.~Season, nrow = 2,  strip.position = "right") 
ordination_RRA
ggsave("Outputs/seasonal_core_species_dietoord_Sept172025.jpeg", ordination_RRA, height = 7.5, width = 9)

```


Can we correlated NMDS1 and NMDS2 with something in the data? i.e., rra willow/
```{r}

### CALCULATING CORRELATION BETWEEN ENVIRONMENTAL COVARS AND AXIS SCORES 
# note that we can't directly convert envfit() vector errors into R2 values for each axis because 
# we did a weighted ordination, not just a regular PCA SO 
# instead we have to calculate r2 values between the abundance of different taxa and NMDS axes manually 
ord_scores <- scores(GP.ord, display = "sites")

otus <- data.frame(t(otu_table(physeq_rra_core)))
tax <-  data.frame(t(tax_table(physeq_rra_core)))
colnames(otus) <- tax["best_id",]
otus_sorted <- otus[,order(colSums(otus), decreasing = TRUE)]
# keep only top 20 columns
otus_sorted <- otus_sorted[,1:20]
otu_names <- colnames(otus_sorted)

# examining the correlation between our ordination axes and our top 20 diet items 
cors <- data.frame(cor(otus[, c(otu_names)], ord_scores, method = "spearman")) 
cors$taxa = rownames(cors)

cors <- pivot_longer(cors, cols = c(1:2)) %>% 
  arrange(., -value)

ggplot(cors, aes(y = reorder(taxa, -value), x = 1, fill = value)) + geom_tile() + 
  scale_fill_viridis_c("Pearson's Correlation") + 
  geom_text(aes(label = round(value, 2))) + theme_bw() + 
  ylab("Taxa Name") + xlab("NMDS Axis") + facet_wrap(.~name, nrow = 1) 

## doing some basic significance testing for our key taxa to get p values
cor.test(ord_scores[, 1], otus[,"Geum triflorum"], method = "spearman")
cor.test(ord_scores[, 1], otus[,"Bistorta vivipara"], method = "spearman")
cor.test(ord_scores[, 1], otus[,"Chamaenerion latifolium"], method = "spearman")
cor.test(ord_scores[, 1], otus[,"Salix spp. "], method = "spearman")
cor.test(ord_scores[, 1], otus[,"Festuca campestris/altaica"], method = "spearman")

cor.test(ord_scores[, 2], otus[,"Salix spp. "], method = "spearman")
cor.test(ord_scores[, 2], otus[,"Rosa acicularis"], method = "spearman")
cor.test(ord_scores[, 2], otus[,"Poaceae (Leymus, Elymus, Agropyron spp.)"], method = "spearman")


```


Making a table with all pianka and mean bray-curtis values for each species-season 

```{r}
### using niche.overlap 
all_spp <- niche.overlap(otu_table(physeq_rra), method = "pianka") # calculate pairwise niche overlap 
# using meandist to get the mean distance 
meandist(all_spp,  sample_data(physeq_rra)$Species)

all_spp <- niche.overlap(otu_table(physeq_rra_core), method = "pianka") # MEAN niche overlap 
meandist(all_spp,  paste(sample_data(physeq_rra_core)$Species, sample_data(physeq_rra_core)$Season))

## bray-curtis 
all_spp_bray <-  vegdist(t(otu_table(physeq_rra)), method = "bray") # calculates bray-curtis dissimilarity 
meandist(all_spp_bray,  sample_data(physeq_rra)$Species)

all_spp_bray <-  vegdist(t(otu_table(physeq_rra_core)), method = "bray") #  MEAN bray-curtis dissimilarity 
meandist(all_spp_bray,  paste(sample_data(physeq_rra_core)$Species, sample_data(physeq_rra_core)$Season))

```



Are bighorn/elk diets in spring/summer different inside/outside of the bison zone? 
```{r}

############### species-biszone interaction
bhs_elk <- subset_samples(physeq_rra_core, (Season == "Spring" | Season == "Summer") & (Species == "Elk" | Species == "BighornSheep"))
table(sample_data(bhs_elk)$Species, sample_data(bhs_elk)$Season)

diet_matrix <- t(otu_table(bhs_elk)) # get our community matrix
covar_table <- data.frame(sample_data(bhs_elk))
in_biszone <- adonis2(diet_matrix ~ Species*Biszone, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
in_biszone


### pooling for ordination 
sp_su <- subset_samples(physeq_rra_core, (Season == "Spring" | Season == "Summer"))
table(sample_data(sp_su)$Species, sample_data(sp_su)$Biszone)

### visualizing using ordination 
GP.ord <- ordinate( t(otu_table(sp_su)) , "NMDS", "bray", trymax=300) # run ordination 

ordination_RRA_dat = plot_ordination(sp_su, GP.ord, color = "Species", shape = "Species", type="samples", justDF = TRUE) %>% 
  group_by(Species, Season) %>% 
  mutate(mean_nmds1 = mean(NMDS1), 
          mean_nmds2 = mean(NMDS2),
          se_nmds1 = sd(NMDS1)/sqrt(n_distinct(SampleID)), 
          se_nmds2 = sd(NMDS2)/sqrt(n_distinct(SampleID)))
ordination_RRA_dat$Species <- factor(ordination_RRA_dat$Species, levels = c("Bison", "BighornSheep", "Elk"), 
                                     labels = c("Bison", "Bighorn sheep", "Elk"))

ordination_RRA_dat$Biszone <- factor(ordination_RRA_dat$Biszone, labels = c("Outside bison area", "Inside bison area"))

ordination_RRA <- ggplot(ordination_RRA_dat, aes(x = NMDS1, y = NMDS2*-1, 
                                                 color = Species, linetype = Species, shape = Species)) +
  geom_point(size = 1.5) + 
  stat_chull(aes(fill = Species, linetype = Species), geom = "polygon", alpha = 0.2, size = 0.8) + 
 # scale_fill_manual("#586028","#898928",  "#00A1B7") + 

#  stat_ellipse(aes(fill = Species, linetype = Species),   size = 0.8) + 

   # scale_color_natparks_d(name = "Banff", direction = -1) + 

  #scale_fill_manual(values = c("firebrick4",  "orange3", "darkorchid4", "orchid1", "gold", "darkorange2")) +   
   # scale_color_manual(values = c("firebrick4",  "orange3", "darkorchid4",  "orchid1",   "gold", "darkorange2")) +
 # scale_shape_manual(values = c(16, 16, 15, 15, 17, 17)) + 
 # scale_linetype_manual(values = c("solid",  "solid", "dashed", "dashed", "dotted", "dotted")) + 
  ylab("NMDS2") + 
     theme_bw() + 
  theme(legend.position ="right") + # c(0.75, 0.23), 
 # geom_text(aes(label = ID_Number), vjust = -0.5) + 
  facet_wrap(.~Biszone, nrow = 2,  strip.position = "right") + 
    scale_fill_manual(values = c("#586028","tan1",  "#00A1B7")) + 
  scale_color_manual(values = c("#586028","tan1",  "#00A1B7")) + 
  scale_shape_manual(values = c(16, 15,  17)) + 
  scale_linetype_manual(values = c("solid", "dashed",  "dotted")) + 
  theme(strip.text = element_text(size = 16))
ordination_RRA

ggsave( "Outputs/inside_outside_diet_comparison.jpeg", plot = ordination_RRA, width = 6, height = 7)


####################### PAIRWISE COMPARISONS 

# elk and bison inside
bis_elk <- subset_samples(physeq_rra_core, Season != "Winter" & (Species == "Bison" | (Species == "Elk" & Biszone == "Y")))
diet_matrix <- t(otu_table(bis_elk)) # get our community matrix
covar_table <- data.frame(sample_data(bis_elk))
bis_elk_inside_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)

# elk and bison outside 
bis_elk <- subset_samples(physeq_rra_core, Season != "Winter" & (Species == "Bison" | (Species == "Elk" & Biszone == "N")))
diet_matrix <- t(otu_table(bis_elk)) # get our community matrix
covar_table <- data.frame(sample_data(bis_elk))
bis_elk_outside_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)

# bhs and bison inside
bis_bhs <- subset_samples(physeq_rra_core, Season != "Winter" & (Species == "Bison" | (Species == "BighornSheep" & Biszone == "Y")))
diet_matrix <- t(otu_table(bis_bhs)) # get our community matrix
covar_table <- data.frame(sample_data(bis_bhs))
bis_bhs_inside_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)

# bhs and bison outside 
bis_bhs <- subset_samples(physeq_rra_core, Season != "Winter" & (Species == "Bison" | (Species == "BighornSheep" & Biszone == "N")))
diet_matrix <- t(otu_table(bis_bhs)) # get our community matrix
covar_table <- data.frame(sample_data(bis_bhs))
bis_bhs_outside_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)

# elk in vs out
elk <- subset_samples(physeq_rra_core, Season != "Winter" & Species == "Elk")
diet_matrix <- t(otu_table(elk)) # get our community matrix
covar_table <- data.frame(sample_data(elk))
elk_in_out <- adonis2(diet_matrix ~ Biszone, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)

# bhs in vs out 
bhs <- subset_samples(physeq_rra_core, Season != "Winter" & Species == "BighornSheep")
diet_matrix <- t(otu_table(bhs)) # get our community matrix
covar_table <- data.frame(sample_data(bhs))
bhs_in_out <- adonis2(diet_matrix ~ Biszone, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)

# vector of p values 
pairwise.seasonal <- data.frame(species = c("bis_elk_inside_pair", "bis_elk_outside_pair", "bis_bhs_inside_pair", "bis_bhs_outside_pair", 
                                   "elk_in_out", "bhs_in_out"),
                      rsq = c(bis_elk_inside_pair$R2[1], 
                                  bis_elk_outside_pair$R2[1], 
                                  bis_bhs_inside_pair$R2[1], 
                                  bis_bhs_outside_pair$R2[1], 
                                  elk_in_out$R2[1], 
                              bhs_in_out$R2[1]
                              ), 
                      pseudoF = c(bis_elk_inside_pair$F[1], 
                                  bis_elk_outside_pair$F[1], 
                                  bis_bhs_inside_pair$F[1], 
                                  bis_bhs_outside_pair$F[1], 
                                  elk_in_out$F[1], 
                              bhs_in_out$F[1]),
                       pval = c( bis_elk_inside_pair$`Pr(>F)`[1], 
                                  bis_elk_outside_pair$`Pr(>F)`[1], 
                                  bis_bhs_inside_pair$`Pr(>F)`[1], 
                                  bis_bhs_outside_pair$`Pr(>F)`[1], 
                                  elk_in_out$`Pr(>F)`[1], 
                              bhs_in_out$`Pr(>F)`[1]))
pairwise.seasonal$holm <-   p.adjust(pairwise.seasonal$pval, "holm", 3) 

write.csv(pairwise.seasonal, "Outputs/biszone_pairwise_comparisons.csv")

```




last but not least, trying out Hannah's PAM algorithm to look at diet clusters 
Use pam() on Bray-Curtis distance matrix to identify dietary clusters using the Partitioning Around Medoids algorithm; use silhouette and elbow methods to identify best cluster number. Sensitivity analyses are conducted in supplement.
```{r}
library(cluster)
library(MASS)

summer <-subset_samples(physeq_rra_core, Season != "Winter")
all_spp_bray <-  vegdist(t(otu_table(summer)), method = "bray", binary = F) #  MEAN bray-curtis dissimilarity 


#bray.dist1228_2 <-metaMDS(all_spp_bray, distance = "bray", k = 4, maxit = 999,  trymax = 200, wascores = TRUE)

bray.dist1228_2_mat <- as.matrix(all_spp_bray)
bray.dist1228_2_mat

#Find optimal number of clusters
asw1228 <- numeric(20)
for (k in 4:length(asw1228)) {
  set.seed(1028)
  asw1228[k] <- cluster::pam(bray.dist1228_2_mat , k) $ silinfo $ avg.width
}
k.best.boot <- which.max(asw1228)
cat("silhouette-optimal number of clusters:", k.best.boot, "\n")	#3 is highest using silhouette method 
plot(asw1228~seq(1,length(asw1228),1),type="l") # 3 is highest also using elbow method 
braydistclust<-vector("list", k.best.boot) #make a list that includes 3 clusters 

#Assign k = 3 from above, and inspect sample clustering
set.seed(100896)
pamNumBoot<-cluster::pam(bray.dist1228_2_mat, k= k.best.boot)
pamNumBoot$clustering
braydistclust<-vector("list", k.best.boot)
braydistclust #this is where samples and their cluster assignments should show up 

#Save copies of each object generated
bray.nmds <- bray.dist1228_2
k.best.bray.sub <- k.best.boot
pamNumBraySub<-pamNumBoot



############ PLOTTING 
### visualizing using ordination 
GP.ord <- ordinate( t(otu_table(summer)) , "NMDS", "bray", trymax=200) # run ordination 
sample_data(summer)$cluster <- as.factor(pamNumBoot$clustering)
ordination_RRA_dat = plot_ordination(summer, GP.ord, color = "Species", shape = "Species", type="samples", justDF = TRUE) 



ggplot(ordination_RRA_dat, aes(x = NMDS1, y = NMDS2*-1, 
                                                 color = cluster, linetype = cluster, shape = cluster)) +
  geom_point(size = 1.5) + 
  stat_chull(aes(fill = cluster, linetype = cluster), geom = "polygon", alpha = 0.2, size = 0.8) + 
  ylab("NMDS2") + 
     theme_bw() + 
  theme(legend.position =  c(0.75, 0.23))+
 # facet_wrap(Season~Species,   strip.position = "right") + 
 #   scale_fill_manual(values = c("#586028","tan1",  "#00A1B7")) + 
#  scale_color_manual(values = c("#586028","tan1",  "#00A1B7")) + 
#  scale_shape_manual(values = c(16, 15,  17)) + 
#  scale_linetype_manual(values = c("solid", "dashed",  "dotted")) + 
  theme(strip.text = element_text(size = 16))

#### looking at clusters 
physeq_df <- psmelt(summer) %>%
   mutate(Sample_description = paste(Sample, Date, General_Loc, sep = ":")) %>%   
  mutate(family_genus_species = paste(family_name, species_name, sep=":"), 
          family_genus = paste(family_name, genus_name, sep = ":")) %>% 
  mutate(Present = ifelse(Abundance > 0, 1, 0))
spp_data <- physeq_df %>%
  group_by(Species, cluster) %>%
  mutate(Seasonal_samples = n_distinct(Sample)) %>%  # number of unique samples in the season 
  group_by(Species, cluster, form, identity) %>% 
  summarise(RRA = sum(Abundance)/unique(Seasonal_samples)*100, 
            Seasonal_samples = unique(Seasonal_samples)) %>% 
    filter(form != "moss") %>% # moss is a miniscule portion, so ditch it! 
  mutate(form= if_else(is.na(form), "other", form)) 
spp_data$form = factor(spp_data$form, levels = c("grass", "non-grass graminoid", "forb",
                                                 "deciduous shrub/tree",
                                                 "coniferous shrub/tree"), 
                                    labels = c("Grass", "Non-grass graminoids", "Forbs", 
                                               "Deciduous shrubs", 
                                               "Coniferous shrubs"))
ggplot(spp_data,
       aes(x = cluster, fill = form, y = RRA)) +
  geom_col(color = "grey20", alpha = 0.8) + 
  scale_x_discrete(expand = c(0, .55)) +
  scale_y_continuous(expand = c(0, 1)) + 
  labs(y = "Relative Read Abundance (%)", x = NULL) +
  theme_bw() +
  #facet_wrap(.~Species) + 
    scale_fill_manual(values = c("gold", "lightyellow2", "darkorchid", "forestgreen", "brown"),
                    na.value = "grey90") + 
  theme(
    axis.ticks = element_blank(),
    legend.position = "top", 
    legend.title = element_blank(),
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()
  ) + 
  guides(fill = guide_legend(reverse = TRUE))


### indicator species analysis for clusters
w_dat <-  summer
otu_matrix <- as.data.frame(otu_table(w_dat))
sample_metadata <- sample_data(w_dat)
groups <-sample_metadata$cluster # our grouping variable = species 

# Add taxonomy to the OTU table
otu_with_tax <- cbind(otu_matrix, tax_table(w_dat))
### GROUP BY mOTU then we can assign stuff post-hoc! 
otu_by_motu <- otu_with_tax %>%
  group_by(identity) %>%
  summarize(across(where(is.numeric), sum, na.rm = TRUE))
otu_by_motu <- t(otu_by_motu) # transpose
colnames(otu_by_motu) <- otu_by_motu[1,] # make labels first row) 
otu_by_motu <- otu_by_motu[-1,] # remove row with labels (character)
otu_by_motu <- apply(otu_by_motu, 2, as.numeric) # convert to numeric! 
otu_by_motu <- otu_by_motu[,colSums(otu_by_motu) > 0.01 ] # remove rare taxa (< 1%)
mOTU_w <- multipatt(otu_by_motu, groups, func = "r.g", control = how(nperm=999))[["sign"]]
mOTU_w$identity = rownames(mOTU_w)
mOTU_w <- left_join(mOTU_w, otu_with_tax[,241:263]) 

mOTU_w <- mOTU_w %>% dplyr::select(s.1, s.2, #s.3, 
                              #     s.4, s.5, s.6, s.7, s.8, s.9, s.10,
                                   index, stat, p.value, identity, family_name, genus_name, best_id, alternative_id)

w_indic <- mOTU_w %>% 
  filter(p.value < 0.05) 
#write.csv(w_indic, "Outputs/WINTER_mOTU_indicspp_allspp.csv")

```


