---
title: "niche overlap"
output: html_document
date: "2024-07-26"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reading in libs
```{r}
library(ggplot2)
library(phyloseq)
library(tidyverse)
library(here)
library(vegan)
library(microViz)
library(ggpubr)
library(spaa)
```

Read in our data 
```{r}
physeq_rra <- readRDS("banff_diet_phyloseq_cleaned_June162026.rds") 

tax_table(physeq_rra)["P114", "best_id"] <- "Rosaceae (Rosa, Rubus, Fragaria spp.)"
tax_table(physeq_rra)["P1", "best_id"] <- "Salicaceae (Salix spp., Populus balsamifera)"
table(tax_table(physeq_rra)[,"form"])
tax_table(physeq_rra)["P114", "form"] <- NA

physeq_rra_core <- subset_samples(physeq_rra, (Species == "Bison" # filter for bison, elk, sheep
                                                      | Species == "BighornSheep" | 
                                                        Species == "Elk"))

sample_data(physeq_rra)$Season = factor(sample_data(physeq_rra)$Season, 
                                         levels = c("Spring", "Summer", "Winter", "Fall"), 
                                         labels = c("Spring", "Summer", "Winter", "Winter"))
sample_data(physeq_rra_core)$Season = factor(sample_data(physeq_rra_core)$Season, 
                                         levels = c("Spring", "Summer", "Winter", "Fall"), 
                                         labels = c("Spring", "Summer", "Winter", "Winter"))


sample_data(physeq_rra)$Species = factor(sample_data(physeq_rra)$Species, 
                                         levels = c("Bison", "BighornSheep",  "Elk", "Goat",  "Deer", "Moose" ), 
                                         labels = c("Bison", "Bighorn Sheep", "Elk",
                                                     "Mountain Goat",  "Deer Spp.", "Moose"))


sample_data(physeq_rra)$Season = factor(sample_data(physeq_rra)$Season, levels = c("Winter", "Summer", "Spring")) 


```



DIFFERENCES IN DIET DUE TO MORPHOLOGY 
```{r}
sample_data(physeq_rra)$feeding_mode <- if_else(sample_data(physeq_rra)$Species == "Bison" | 
                                                  sample_data(physeq_rra)$Species == "Bighorn Sheep", "Grazer", 
                                if_else(sample_data(physeq_rra)$Species== "Deer Spp.", "Browser", "Mixed"))

feeding_mode_dat <- subset_samples(physeq_rra, (Species != "Moose" & Species != "Mountain Goat"))



# group species by functional feeding more and see if permanova differentiates them... use spicy pairwise comparisons too!! 
diet_matrix <- t(otu_table(feeding_mode_dat)) # get our community matrix
covar_table <- data.frame(sample_data(feeding_mode_dat)) 
# run adonis test... community data matrix ~ covariate (rows in community matrix match rows in covar data frame)
table(covar_table$feeding_mode)
ffmode_mod <- adonis2(diet_matrix ~ feeding_mode, data = covar_table, permutations = 999, 
                method = "bray",
    parallel = 15, na.action = na.fail, by = "terms")

ffmode_mod


# group species by functional feeding more and see if permanova differentiates them... use spicy pairwise comparisons too!! 
graz_browse <- subset_samples(feeding_mode_dat, (feeding_mode == "Grazer" # filter for bison, elk, sheep
                                                      | feeding_mode == "Browser"))

diet_matrix <- t(otu_table(graz_browse)) # get our community matrix
covar_table <- data.frame(sample_data(graz_browse)) 
# run adonis test... community data matrix ~ covariate (rows in community matrix match rows in covar data frame)
table(covar_table$feeding_mode)
ffmode_mod <- adonis2(diet_matrix ~ feeding_mode, data = covar_table, permutations = 999, 
                method = "bray",
    parallel = 15, na.action = na.fail, by = "terms")

ffmode_mod

## graz mixed! 
graz_browse <- subset_samples(feeding_mode_dat, (feeding_mode == "Grazer" # filter for bison, elk, sheep
                                                      | feeding_mode == "Mixed"))

diet_matrix <- t(otu_table(graz_browse)) # get our community matrix
covar_table <- data.frame(sample_data(graz_browse)) 
# run adonis test... community data matrix ~ covariate (rows in community matrix match rows in covar data frame)
table(covar_table$feeding_mode)
ffmode_mod <- adonis2(diet_matrix ~ feeding_mode, data = covar_table, permutations = 999, 
                method = "bray",
    parallel = 15, na.action = na.fail, by = "terms")

ffmode_mod
p.adjust(ffmode_mod$`Pr(>F)`, "holm", n = 3) 


graz_browse <- subset_samples(feeding_mode_dat, (feeding_mode == "Browser" # filter for bison, elk, sheep
                                                      | feeding_mode == "Mixed"))

diet_matrix <- t(otu_table(graz_browse)) # get our community matrix
covar_table <- data.frame(sample_data(graz_browse)) 
# run adonis test... community data matrix ~ covariate (rows in community matrix match rows in covar data frame)
table(covar_table$feeding_mode)
ffmode_mod <- adonis2(diet_matrix ~ feeding_mode, data = covar_table, permutations = 999, 
                method = "bray",
    parallel = 15, na.action = na.fail, by = "terms")

ffmode_mod

p.adjust(ffmode_mod$`Pr(>F)`, "holm", n = 3) 
```

BODY SIZE DIFFERENCES? 
```{r}
######## PAIRWISE COMPARISONS 

# bison sheep 
bis_bhs <- subset_samples(physeq_rra, Species == "Bison" | Species == "Bighorn Sheep")
diet_matrix <- t(otu_table(bis_bhs)) # get our community matrix
covar_table <- data.frame(sample_data(bis_bhs))
bis_bhs_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)

# bison elk 
bis_elk <- subset_samples(physeq_rra, Species == "Bison" | (Species == "Elk" & Long < -115.4105))
diet_matrix <- t(otu_table(bis_elk)) # get our community matrix
covar_table <- data.frame(sample_data(bis_elk))
bis_elk_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)

# bison deer
bis_deer  <- subset_samples(physeq_rra, Species == "Bison" | Species == "Deer Spp.")
diet_matrix <- t(otu_table(bis_deer)) # get our community matrix
covar_table <- data.frame(sample_data(bis_deer))
bis_deer_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bis_deer_pair


# bighorn slk 
dat  <- subset_samples(physeq_rra, Species == "Bighorn Sheep" | Species == "Elk")
diet_matrix <- t(otu_table(dat)) # get our community matrix
covar_table <- data.frame(sample_data(dat))
bhs_elk_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bhs_elk_pair


# bighorn deer 
dat  <- subset_samples(physeq_rra, Species == "Bighorn Sheep" | Species == "Deer Spp.")
diet_matrix <- t(otu_table(dat)) # get our community matrix
covar_table <- data.frame(sample_data(dat))
bhs_deer_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bhs_deer_pair


# elk - deer 
dat  <- subset_samples(physeq_rra, Species == "Elk" | Species == "Deer Spp.")
diet_matrix <- t(otu_table(dat)) # get our community matrix
covar_table <- data.frame(sample_data(dat))
elk_deer_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
elk_deer_pair



# get our adjusted p values
pairwise.species <- data.frame(species = c("bis_bhs", "bis_elk", "bis_deer", 
                                           "bhs_elk", "bhs_deer",
                                           "elk_deer"), 
                              rsq = c(bis_bhs_pair$R2[1],
                                        bis_elk_pair$R2[1], 
                                        bis_deer_pair$R2[1],
                                      #
                                      bhs_elk_pair$R2[1],
                                        bhs_deer_pair$R2[1],
                                      #
                                        elk_deer_pair$R2[1]),
                                      #
                                  pseudoF = c(bis_bhs_pair$F[1],
                                        bis_elk_pair$F[1], 
                                        bis_deer_pair$F[1],
                                      #
                                      bhs_elk_pair$F[1],
                                        bhs_deer_pair$F[1],
                                      #
                                        elk_deer_pair$F[1]),
                                  pval = c(bis_bhs_pair$`Pr(>F)`[1],
                                        bis_elk_pair$`Pr(>F)`[1], 
                                        bis_deer_pair$`Pr(>F)`[1],
                                      #
                                      bhs_elk_pair$`Pr(>F)`[1],
                                        bhs_deer_pair$`Pr(>F)`[1],
                                      #
                                        elk_deer_pair$`Pr(>F)`[1]),
                                  
                                body_diff = c(550, 375, 545, 175, 5, 170))


pairwise.species$holm <-   p.adjust(pairwise.species$pval, "holm", n = 6 ) 


write.csv(pairwise.species, "Outputs/pairwise_comparisons_across_all_species_july2026.csv")


pairwise.species <- read.csv("Outputs/pairwise_comparisons_across_all_species_dec082024.csv")

####### BODY SIZE DIFFERENCES USING MANTEL TEST! 
# Split species names
pairwise.species_split <- tidyr::separate(pairwise.species, species, into=c("sp1","sp2"), sep="_") 
pairwise.species_split <- rbind(pairwise.species_split, data.frame(sp1 = c("bis", "bhs", "elk", "deer"),sp2= c("bis", "bhs", "elk", "deer"), 
                                                                   rsq = NA, pseudoF = NA, pval = NA, body_diff  = NA,holm = NA))

# Make square matrix of species × species
species <- unique(c(pairwise.species_split$sp1, pairwise.species_split$sp2))
rsq_mat <- matrix(NA, nrow=length(species), ncol=length(species),
                  dimnames=list(species, species))
size_mat <- matrix(NA, nrow=length(species), ncol=length(species),
                  dimnames=list(species, species))
# Fill matrix
for(i in 1:nrow(pairwise.species_split)) {
  rsq_mat[pairwise.species_split$sp1[i], pairwise.species_split$sp2[i]] <- pairwise.species_split$rsq[i]
  rsq_mat[pairwise.species_split$sp2[i], pairwise.species_split$sp1[i]] <- pairwise.species_split$rsq[i]  # make symmetric
}
diag(rsq_mat) <- 0
# Fill matrix
for(i in 1:nrow(pairwise.species_split)) {
  size_mat[pairwise.species_split$sp1[i], pairwise.species_split$sp2[i]] <- pairwise.species_split$body_diff[i]
  size_mat[pairwise.species_split$sp2[i], pairwise.species_split$sp1[i]] <- pairwise.species_split$body_diff[i]  # make symmetric
}
diag(size_mat) <- 0

# converting to distance objects!
diss_dist <- as.dist(rsq_mat)
size_dist <- as.dist(size_mat)

# mantel test! 
mantel_res <- mantel(diss_dist, size_dist, method = "pearson", permutations = 999)
print(mantel_res)


```



Adonis test to examine if there are seasonal differences in diets of elk/bison/sheep. Followed by some nice pairwise comaprisons
```{r}
## all species-season! 
diet_matrix <- t(otu_table(physeq_rra_core)) # get our community matrix
covar_table <- data.frame(sample_data(physeq_rra_core))
class(covar_table)
# run adonis test... community data matrix ~ covariate (rows in community matrix match rows in covar data frame)
season_mod <- adonis2(diet_matrix ~ Species*Season, data = covar_table, permutations = 999, 
                method = "bray",
    parallel = 15, na.action = na.fail, by = "terms")

season_mod

#rm(season_mod)

# WITHIN season differences 

sp <- subset_samples(physeq_rra_core, Season == "Spring")
diet_matrix <- t(otu_table(sp)) # get our community matrix
covar_table <- data.frame(sample_data(sp))
sp_adonis <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail, by = "terms")
sp_adonis

su <- subset_samples(physeq_rra_core, Season == "Summer")
diet_matrix <- t(otu_table(su)) # get our community matrix
covar_table <- data.frame(sample_data(su))
su_adonis <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail, by = "terms")
su_adonis




w <- subset_samples(physeq_rra_core, Season == "Winter" | Season == "Fall")
diet_matrix <- t(otu_table(w)) # get our community matrix
covar_table <- data.frame(sample_data(w))
w_adonis <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
w_adonis


############# PAIRWISE ADONIS ##########################
# use adonis to compare species-season.... 
#################### SEASON SPECIFIC COMPARISONS 

# elk and bison winter
bis_elk <- subset_samples(physeq_rra_core, Season == "Winter" & (Species == "Bison" | Species == "Elk"))
diet_matrix <- t(otu_table(bis_elk)) # get our community matrix
covar_table <- data.frame(sample_data(bis_elk))
bis_elk_w_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bis_elk_w_pair

# bhs and bison winter
bis_bhs <- subset_samples(physeq_rra_core, Season == "Winter" & (Species == "Bison" | Species == "BighornSheep"))
diet_matrix <- t(otu_table(bis_bhs)) # get our community matrix
covar_table <- data.frame(sample_data(bis_bhs))
bis_bhs_w_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bis_bhs_w_pair

# elk and bison spring
bis_elk <- subset_samples(physeq_rra_core, Season == "Spring" & (Species == "Bison" | Species == "Elk"))
diet_matrix <- t(otu_table(bis_elk)) # get our community matrix
covar_table <- data.frame(sample_data(bis_elk))
bis_elk_sp_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bis_elk_sp_pair

# bhs and bison spring
bis_bhs <- subset_samples(physeq_rra_core, Season == "Spring" & (Species == "Bison" | Species == "BighornSheep"))
diet_matrix <- t(otu_table(bis_bhs)) # get our community matrix
covar_table <- data.frame(sample_data(bis_bhs))
bis_bhs_sp_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bis_bhs_sp_pair

# elk and bison summer
bis_elk <- subset_samples(physeq_rra_core, Season == "Summer" & (Species == "Bison" | Species == "Elk"))
diet_matrix <- t(otu_table(bis_elk)) # get our community matrix
covar_table <- data.frame(sample_data(bis_elk))
bis_elk_su_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bis_elk_su_pair

# bhs and bison SUMMER
bis_bhs <- subset_samples(physeq_rra_core, Season == "Summer" & (Species == "Bison" | Species == "BighornSheep"))
diet_matrix <- t(otu_table(bis_bhs)) # get our community matrix
covar_table <- data.frame(sample_data(bis_bhs))
bis_bhs_su_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bis_bhs_su_pair

# bhs and elk spring
dat <- subset_samples(physeq_rra_core, Season == "Spring" & (Species == "Elk" | Species == "BighornSheep"))
diet_matrix <- t(otu_table(dat)) # get our community matrix
covar_table <- data.frame(sample_data(dat))
bhs_elk_sp_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bhs_elk_sp_pair

# bjs and elk summer
dat <- subset_samples(physeq_rra_core, Season == "Summer" & (Species == "Elk" | Species == "BighornSheep"))
diet_matrix <- t(otu_table(dat)) # get our community matrix
covar_table <- data.frame(sample_data(dat))
bhs_elk_su_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bhs_elk_su_pair


# bjs and elk winter
dat <- subset_samples(physeq_rra_core, Season == "Winter" & (Species == "Elk" | Species == "BighornSheep"))
diet_matrix <- t(otu_table(dat)) # get our community matrix
covar_table <- data.frame(sample_data(dat))
bhs_elk_w_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail)
bhs_elk_w_pair


# vector of p values 
pairwise.seasonal <- data.frame(species = c("bis_elk_w_pair", "bis_bhs_w_pair", "bis_elk_sp_pair", "bis_bhs_sp_pair", 
                                   "bis_elk_su_pair", "bis_bhs_su_pair", 
                                   "bhs_elk_sp", "bhs_elk_su", "bhs_elk_w"),
                      rsq = c(bis_elk_w_pair$R2[1], 
                                  bis_bhs_w_pair$R2[1], 
                                  bis_elk_sp_pair$R2[1], 
                                  bis_bhs_sp_pair$R2[1], 
                                  bis_elk_su_pair$R2[1], 
                                  bis_bhs_su_pair$R2[1],
                              bhs_elk_sp_pair$R2[1],
                              bhs_elk_su_pair$R2[1],
                              bhs_elk_w_pair$R2[1]
                              ), 
                      pseudoF = c(bis_elk_w_pair$F[1], 
                                  bis_bhs_w_pair$F[1], 
                                  bis_elk_sp_pair$F[1], 
                                  bis_bhs_sp_pair$F[1], 
                                  bis_elk_su_pair$F[1], 
                                  bis_bhs_su_pair$F[1], 
                                
                              bhs_elk_sp_pair$F[1],
                              bhs_elk_su_pair$F[1],
                              bhs_elk_w_pair$F[1]),
                       pval = c( bis_elk_w_pair$`Pr(>F)`[1], 
                                  bis_bhs_w_pair$`Pr(>F)`[1], 
                                  bis_elk_sp_pair$`Pr(>F)`[1], 
                                  bis_bhs_sp_pair$`Pr(>F)`[1], 
                                  bis_elk_su_pair$`Pr(>F)`[1], 
                                  bis_bhs_su_pair$`Pr(>F)`[1], 
                                 
                              bhs_elk_sp_pair$`Pr(>F)`[1],
                              bhs_elk_su_pair$`Pr(>F)`[1],
                              bhs_elk_w_pair$`Pr(>F)`[1]))
pairwise.seasonal$holm <-   p.adjust(pairwise.seasonal$pval, "holm", 9) 

write.csv(pairwise.seasonal, "Outputs/pairwise_comparisons_across_season_july2026.csv")
```



Visualizing using NMDS 
Seasonal differences visualized (bison, sheep, elk) 
```{r}

################## Removing deer, moose, and goats (from the figure in main text) 
GP.ord <- ordinate(physeq_rra_core , "NMDS", "bray", trymax=300) # run ordination 


### CALCULATING CORRELATION BETWEEN ENVIRONMENTAL COVARS AND AXIS SCORES 
# note that we can't directly convert envfit() vector errors into R2 values for each axis because 
# we did a weighted ordination, not just a regular PCA SO 
# instead we have to calculate r2 values between the abundance of different taxa and NMDS axes manually 
ord_scores <- scores(GP.ord, display = "sites")

otus <- data.frame(t(otu_table(physeq_rra_core)))
tax <-  data.frame(t(tax_table(physeq_rra_core)))
colnames(otus) <- tax["best_id",]
otus_sorted <- otus[,order(colSums(otus), decreasing = TRUE)]
# keep only top 20 columns
#### WHAT IF WE JUST PICK OUR INDICATOR SPECIES? 

# names based on figure 5 (indicator species)
otus_sorted <- otus_sorted[,colnames(otus_sorted) %in% c("Geum triflorum",
                                                         "Salicaceae (Salix spp., Populus balsamifera)",
                                                         "Poaceae (Leymus, Elymus, Agropyron spp.)",
                                                         "Festuca campestris/altaica", 
                                                         "Rosaceae (Rosa, Rubus, Fragaria spp.)", 
                                                         "Chamaenerion angustifolium", 
                                                         "Potentilla gracilis", 
                                                         "Chamaenerion latifolium", 
                                                         "Alsineae (Cerastium, Stellaria spp.)", 
                                                         "Bistorta vivipara", 
                                                         "Carex nardina", 
                                                         "Picea spp. ")]
otu_names <- colnames(otus_sorted)


# examining the correlation between our ordination axes and our top 20 diet items 
# which plant taxa? 
library(psych)
cor_test_results <- corr.test(otus[, c(otu_names)], ord_scores, method = "spearman", adjust = "holm")

# extract the coefficients (r) and convert to a dataframe
r_df <- as.data.frame(cor_test_results$r) %>% 
  rownames_to_column(var = "taxa") %>% 
  rename(r_NMDS1 = NMDS1, r_NMDS2 = NMDS2)

# extract the adjusted p-values and convert to a dataframe
p_df <- as.data.frame(cor_test_results$p) %>% 
  rownames_to_column(var = "taxa") %>% 
  rename(p_NMDS1 = NMDS1, p_NMDS2 = NMDS2)

strong_drivers <-inner_join(r_df, p_df, by = "taxa")

#### PLOTS 
# first, just basic plot! 
strong_drivers %>% 
  arrange(-abs(r_NMDS1))

strong_drivers %>% 
  arrange(-abs(r_NMDS2))

strong_drivers <- strong_drivers %>%
  mutate(
    # Create numbers 1, 2, 3... based on row order
    taxa_num = as.character(row_number()), 
    allszn = as.factor("All seasons") ) # Forces ggplot to only draw this on the first facet


write.csv(strong_drivers, "Outputs/nmds_correlations.csv")

## getting data for plotting 
ordination_RRA_dat = plot_ordination(physeq_rra_core, GP.ord, color = "Species", shape = "Species", type="samples", justDF = TRUE) %>% 
  group_by(Species, Season) %>% 
  mutate(mean_nmds1 = mean(NMDS1), 
          mean_nmds2 = mean(NMDS2),
          se_nmds1 = sd(NMDS1)/sqrt(n_distinct(SampleID)), 
          se_nmds2 = sd(NMDS2)/sqrt(n_distinct(SampleID)))

ordination_RRA_dat$Species = factor(ordination_RRA_dat$Species, levels = c("Bison", "BighornSheep", "Elk"), 
                                    labels = c("Bison", "Bighorn sheep", "Elk"))
table(ordination_RRA_dat$Species)

ordination_RRA_dat$Season = factor(ordination_RRA_dat$Season, levels = c("Winter", "Summer", "Spring"))
# We're going to make a figure with 4 panels, 1 is all species/seasons, then the other 3 are seasonal diets 

ordination_RRA_dat$allszn <- "All seasons"
ordination_rra_temp <-ordination_RRA_dat %>% mutate(allszn = Season)

ordination_RRA_dat2 <- rbind(ordination_RRA_dat, ordination_rra_temp)

ordination_RRA_dat2$allszn <- factor(ordination_RRA_dat2$allszn, levels = c("All seasons", "Winter", "Spring", "Summer"))

plot <- ggplot() + 
  geom_point(data = ordination_RRA_dat2, aes(x = NMDS1, y = NMDS2, 
                                             color = Species, linetype = Species, 
                                             shape = Species, fill = Species), size = 1) + 
  stat_ellipse(data = ordination_RRA_dat2,  level = 0.3935, type = "norm",
               aes(x = NMDS1, y = NMDS2, 
                                                color = Species, linetype = Species, 
                                                shape = Species, fill = Species), geom = "polygon", alpha = 0.2, size = 0.6) + 
  
  scale_fill_manual(values = c("#586028","tan1",  "#00A1B7",  "grey30",  "grey50",  "grey80")) + 
  scale_color_manual(values = c("#586028","tan1",  "#00A1B7", "grey30",  "grey50",  "grey80")) + 
  scale_shape_manual(values = c(16, 15,  17, 16, 16, 16)) + 
  scale_linetype_manual(values = c("solid", "dashed",  "dotted", "dotted", "dotted", "dotted")) + 
  ylab("NMDS2") + 
  theme_bw() + 
  theme(legend.position = "right", 
        legend.title = element_blank(),
        strip.text = element_text(size = 12),   
        axis.ticks = element_blank(), axis.text = element_blank(),
        axis.title = element_text(size = 12),
        legend.key = element_blank(), 
        panel.grid.major = element_line(color = "grey92", linewidth = 0.3),
        panel.grid.minor = element_line(color = "grey96", linewidth = 0.3)) + 

  facet_wrap(.~allszn, nrow = 2, strip.position = "top") + 
  
  # 1. Scaled arrows
  geom_segment(data = strong_drivers, 
               aes(x = 0, y = 0, 
                   xend = r_NMDS1 * 2, 
                   yend = r_NMDS2 * 2),
               arrow = arrow(length = unit(0.1, "cm")), color = "black", size = 0.5) +

  # 2. FIXED: Uses ggrepel to push numbers cleanly off the tips without leader lines
geom_text(data = strong_drivers, 
                           aes(x = r_NMDS1 * 2.2, 
                               y = r_NMDS2 * 2.2, 
                               label = taxa_num, 
                               alpha = taxa), 
                           color = "black", 
                           fontface = "bold",
                           size = 3.8,
                           box.padding = 0.05, 
                   segment.color = "black", segment.size = 0.5) +
  scale_alpha_manual(name = "Taxa Drivers", 
                     breaks = strong_drivers$taxa,
                     values = rep(1, nrow(strong_drivers))) +
  guides(
    color = guide_legend(title = NULL, order = 1),
    shape = guide_legend(title = NULL, order = 1),
    linetype = guide_legend(title = NULL, order = 1),
    fill = guide_legend(title = NULL, order = 1),
    
    alpha = guide_legend(
      title = "Taxa Drivers", 
      ncol = 1,                                                
      label.theme = element_text(face = "italic"),         
      order = 2,                                         
      override.aes = list(
        label = strong_drivers$taxa_num,  
        fontface = "bold"                                  
      )
    )
  )

 plot

ggsave(plot, filename ="Outputs/NMDS_Aug122026.jpeg", height = 6, width = 10, dpi = 300)


```




Making a table with all pianka and mean bray-curtis values for each species-season 

```{r}
### using niche.overlap 
all_spp <- niche.overlap(otu_table(physeq_rra), method = "pianka") # calculate pairwise niche overlap 
# using meandist to get the mean distance 
meandist(all_spp,  sample_data(physeq_rra)$Species)

all_spp <- niche.overlap(otu_table(physeq_rra_core), method = "pianka") # MEAN niche overlap 
meandist(all_spp,  paste(sample_data(physeq_rra_core)$Species, sample_data(physeq_rra_core)$Season))

## bray-curtis 
all_spp_bray <-  vegdist(t(otu_table(physeq_rra)), method = "bray") # calculates bray-curtis dissimilarity 
meandist(all_spp_bray,  sample_data(physeq_rra)$Species)

all_spp_bray <-  vegdist(t(otu_table(physeq_rra_core)), method = "bray") #  MEAN bray-curtis dissimilarity 
meandist(all_spp_bray,  paste(sample_data(physeq_rra_core)$Species, sample_data(physeq_rra_core)$Season))

```






Are bighorn/elk diets in spring/summer different inside/outside of the bison zone? 
```{r}

############### species-biszone interaction
bhs_elk <- subset_samples(physeq_rra_core, (Season == "Spring" | Season == "Summer") & (Species == "Elk" | Species == "BighornSheep"))
table(sample_data(bhs_elk)$Species, sample_data(bhs_elk)$Season)

diet_matrix <- t(otu_table(bhs_elk)) # get our community matrix
covar_table <- data.frame(sample_data(bhs_elk))
in_biszone <- adonis2(diet_matrix ~ Species*Biszone, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail, by = "terms")
in_biszone


### pooling for ordination 
sp_su <- subset_samples(physeq_rra_core, (Season == "Spring" | Season == "Summer"))
table(sample_data(sp_su)$Species, sample_data(sp_su)$Biszone)

### visualizing using ordination 
GP.ord <- ordinate( t(otu_table(sp_su)) , "NMDS", "bray", trymax=300) # run ordination 

ordination_RRA_dat = plot_ordination(sp_su, GP.ord, color = "Species", shape = "Species", type="samples", justDF = TRUE) %>% 
  group_by(Species, Season) %>% 
  mutate(mean_nmds1 = mean(NMDS1), 
          mean_nmds2 = mean(NMDS2),
          se_nmds1 = sd(NMDS1)/sqrt(n_distinct(SampleID)), 
          se_nmds2 = sd(NMDS2)/sqrt(n_distinct(SampleID)))
ordination_RRA_dat$Species <- factor(ordination_RRA_dat$Species, levels = c("Bison", "BighornSheep", "Elk"), 
                                     labels = c("Bison", "Bighorn sheep", "Elk"))

ordination_RRA_dat$Biszone <- factor(ordination_RRA_dat$Biszone, labels = c("Outside bison zone", "Inside bison zone"))

ordination_RRA <- ggplot(ordination_RRA_dat, aes(x = NMDS1, y = NMDS2*-1, 
                                                 color = Species, linetype = Species, shape = Species, fill = Species)) +
  geom_point(size = 2) + 
  stat_ellipse(aes(fill = Species, linetype = Species), type = "norm", level = 0.3875, geom = "polygon", alpha = 0.2, size = 0.8) + 
  ylab("NMDS2") + 
     theme_bw() + 
  theme(legend.position = "top") + 
 # geom_text(aes(label = ID_Number), vjust = -0.5) + 
  facet_wrap(.~Biszone, nrow = 2,  strip.position = "right") + 
    scale_fill_manual(values = c("#586028","tan1",  "#00A1B7")) + 
  scale_color_manual(values = c("#586028","tan1",  "#00A1B7")) + 
  scale_shape_manual(values = c(16, 15,  17)) + 
  scale_linetype_manual(values = c("solid", "dashed",  "dotted")) + 
  theme(strip.text = element_text(size = 16), 
        axis.title = element_text(size = 16),
        legend.title = element_blank(),
        axis.text = element_blank(), axis.ticks = element_blank())
ordination_RRA

ggsave( "Outputs/inside_outside_diet_comparison_Aug122026.jpeg", plot = ordination_RRA, width = 5, height = 8.5)


####################### PAIRWISE COMPARISONS 

# elk and bison inside
bis_elk <- subset_samples(physeq_rra_core, Season != "Winter" & (Species == "Bison" | (Species == "Elk" & Biszone == "Y")))
diet_matrix <- t(otu_table(bis_elk)) # get our community matrix
covar_table <- data.frame(sample_data(bis_elk))
bis_elk_inside_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail, by = "terms")

# elk and bison outside 
bis_elk <- subset_samples(physeq_rra_core, Season != "Winter" & (Species == "Bison" | (Species == "Elk" & Biszone == "N")))
diet_matrix <- t(otu_table(bis_elk)) # get our community matrix
covar_table <- data.frame(sample_data(bis_elk))
bis_elk_outside_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail, by = "terms")

# bhs and bison inside
bis_bhs <- subset_samples(physeq_rra_core, Season != "Winter" & (Species == "Bison" | (Species == "BighornSheep" & Biszone == "Y")))
diet_matrix <- t(otu_table(bis_bhs)) # get our community matrix
covar_table <- data.frame(sample_data(bis_bhs))
bis_bhs_inside_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail, by = "terms")

# bhs and bison outside 
bis_bhs <- subset_samples(physeq_rra_core, Season != "Winter" & (Species == "Bison" | (Species == "BighornSheep" & Biszone == "N")))
diet_matrix <- t(otu_table(bis_bhs)) # get our community matrix
covar_table <- data.frame(sample_data(bis_bhs))
bis_bhs_outside_pair <- adonis2(diet_matrix ~ Species, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail, by = "terms")

# elk in vs out
elk <- subset_samples(physeq_rra_core, Season != "Winter" & Species == "Elk")
diet_matrix <- t(otu_table(elk)) # get our community matrix
covar_table <- data.frame(sample_data(elk))
elk_in_out <- adonis2(diet_matrix ~ Biszone, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail, by = "terms")

# bhs in vs out 
bhs <- subset_samples(physeq_rra_core, Season != "Winter" & Species == "BighornSheep")
diet_matrix <- t(otu_table(bhs)) # get our community matrix
covar_table <- data.frame(sample_data(bhs))
bhs_in_out <- adonis2(diet_matrix ~ Biszone, data = covar_table, permutations = 999, 
                method = "bray", parallel = 15, na.action = na.fail, by = "terms")

# vector of p values 
pairwise.seasonal <- data.frame(species = c("bis_elk_inside_pair", "bis_elk_outside_pair", "bis_bhs_inside_pair", "bis_bhs_outside_pair", 
                                   "elk_in_out", "bhs_in_out"),
                      rsq = c(bis_elk_inside_pair$R2[1], 
                                  bis_elk_outside_pair$R2[1], 
                                  bis_bhs_inside_pair$R2[1], 
                                  bis_bhs_outside_pair$R2[1], 
                                  elk_in_out$R2[1], 
                              bhs_in_out$R2[1]
                              ), 
                      pseudoF = c(bis_elk_inside_pair$F[1], 
                                  bis_elk_outside_pair$F[1], 
                                  bis_bhs_inside_pair$F[1], 
                                  bis_bhs_outside_pair$F[1], 
                                  elk_in_out$F[1], 
                              bhs_in_out$F[1]),
                       pval = c( bis_elk_inside_pair$`Pr(>F)`[1], 
                                  bis_elk_outside_pair$`Pr(>F)`[1], 
                                  bis_bhs_inside_pair$`Pr(>F)`[1], 
                                  bis_bhs_outside_pair$`Pr(>F)`[1], 
                                  elk_in_out$`Pr(>F)`[1], 
                              bhs_in_out$`Pr(>F)`[1]))
pairwise.seasonal$holm <-   p.adjust(pairwise.seasonal$pval, "holm", 6) 

write.csv(pairwise.seasonal, "Outputs/biszone_pairwise_comparisons.csv")


### I want to make some simple bar plots to show dietary proportions of elk and bighorn sheep in and out of the bison zone, for comparison's sake 
physeq_comparison <- psmelt(physeq_rra_core)  %>%
    mutate(Season = if_else(Season == "Fall", "Winter", Season)) %>% 
  filter(Season != "Winter") %>% 
  group_by(Species, Biszone) %>%
  mutate(Seasonal_samples = n_distinct(Sample)) %>%
  group_by(Species,Biszone, form, identity) %>%
  summarize(RRA = sum(Abundance)/unique(Seasonal_samples), 
             Seasonal_samples = unique(Seasonal_samples))%>% 
  mutate(form= if_else(is.na(form) | form == "moss", "other", form))


physeq_comparison$form = factor(physeq_comparison$form, levels = c("grass", "non-grass graminoid", "forb", 
                                                                     "deciduous shrub/tree",
                                                                     "coniferous shrub/tree", "other"), 
                                   labels = c("Grass", "Non-grass graminoids", "Forbs", 
                                               "Deciduous shrubs", 
                                               "Coniferous shrubs", "Other/unknown"))


physeq_comparison <- physeq_comparison  %>% 
  filter(!is.na(form))
physeq_comparison$Species = factor(physeq_comparison$Species, 
                                       levels = c("Bison", "BighornSheep", "Elk"), 
                                       labels = c("Bison", "Bighorn Sheep", "Elk"))

plot <- ggplot(physeq_comparison[physeq_comparison$Species != "Bison",],
       aes(x = Species, fill = form, y = RRA * 100)) +
  geom_col(color = "grey20", alpha = 0.8) + 
  scale_x_discrete(expand = c(0, 0.55)) + 
  scale_y_continuous(expand = c(0, 1), position = "right") +
  theme_bw() +
  facet_wrap(.~Biszone, ncol = 1) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.ticks = element_blank(),
    axis.title.x = element_blank(),
    panel.spacing = unit(0.5, "cm"),
    strip.text = element_blank(),
    axis.text.y = element_text(size = 10), 
    axis.title.y = element_text(size = 15),
    legend.position = "top", 
    legend.title = element_blank(), 
    legend.text = element_text(size = 12),
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()
  ) + 
  scale_fill_manual(values = c("gold", "lightyellow3", "darkorchid", "forestgreen", "brown", "grey98")) + 
  
  # MODIFIED: Forces the fill legend into 2 columns
  guides(fill = guide_legend(ncol = 2)) + 
  
  ylab("Relative Read Abundance (%)")

plot

ggsave( "Outputs/diet_comp_inside_outside_diet_comparison.jpeg", plot = plot, width = 4.5, height = 9)



```

